---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# volcalc

<!-- badges: start -->
<!-- badges: end -->

The goal of volcalc is to automate calculating estimates of volatility for chemical compounds.
It also provides the ability to calculate the volatility of all compounds in a metabolic pathway.
volcalc leverages the [ChemmineR](https://www.bioconductor.org/packages/devel/bioc/vignettes/ChemmineR/inst/doc/ChemmineR.html) (Cao et al 2008) package to query structural information about chemical compounds, and the [KEGGREST](https://bioconductor.org/packages/release/bioc/html/KEGGREST.html) (Tenenbaum et al 2021) package to query the [KEGG database](https://www.genome.jp/kegg/) for metabolic pathways.
Volatility calculation uses the SIMPOL  method (Prankow and Asher, 2008), and is only currently applicable to compounds and pathways in KEGG.

## Installing and loading package

### Gaining Access

<!-- this section can be deleted after the repository is made public--> 

Because the package's GitHub repository is private, you'll need to use an authentication method. 

- Option 1: If you already have an ssh key, you can register it with GitHub at https://github.com/settings/keys; as described in in ['Connecting to GitHub with SSH'](https://docs.github.com/en/authentication/connecting-to-github-with-ssh). 


- Option 2: create a personal access token to install it. Get token by going to [https://github.com/settings/tokens](https://github.com/settings/tokens). Hit "Generate new token" button, give the token a name like "volcalc package", and make sure there is a check by the "repo" option before hitting green button at bottom of page. Copy the token, which is a random string of numbers and letters, on the next page and insert into the R command below. Do not share your token with anyone! 

### Installing the package

You can install volcalc from [the GitHub repository](https://github.com/Meredith-Lab/volcalc) with:

``` r
# install.packages("devtools")
devtools::install_github("Meredith-Lab/volcalc", auth_token = "[token]")
```

You might be prompted to install some additional R packages, follow the instructions provided to do so. 

And then use the package with: 

```{r}
library(volcalc)
```

## USE

### Estimate volatility of a single compound

We will start with a basic example that shows you how to estimate the volatility of a single compound, *beta-2,3,4,5,6-Pentachlorocyclohexanol*. 
You can search for the KEGG compound identifier in the KEGG database at https://www.genome.jp/kegg/. 
For this compound, the identifier is *C16181*, and more information about this compound can be found at https://www.genome.jp/entry/C16181. 
We can see that it is part of two molecular pathways; for this example we will use the first, *map00361*, below. 

#### Single function approach

Here, we calcuate the volatility of the compound 	_beta-2,3,4,5,6-Pentachlorocyclohexanol_.

```{r}
calc_vol("map00361", "data", compound_id = "C16181")
```

This returns a dataframe with columns containing general information about the compound, the compound's calculated volatility, and its corresponding volatility category. The functional group counts underlying the volatility can be additionally returned with `return_fx_groups = TRUE`, and the intermediate calculation steps with `return_calc_steps = TRUE`. A list of all possible dataframe columns is included below. 

The compound can alternatively be specified with its chemical formula using the `compound_formula` argument instead of `compound_id` as in the example.

#### Multiple function approach

This breaks the steps done by `calc_vol` into three parts: 1) download the compound's .mol file from KEGG, 2) count occurrences of different functional groups, and 3) estimate volatility. This calculation uses the SIMPOL approach (Prankow and Asher, 2008).

```{r}
save_compound_mol("map00361", "data/", compound_id = "C16181")
example_compound_fx_groups <- get_fx_groups("C16181", "map00361", "data/")
example_compound_vol <- calc_vol("map00361", "data/", compound_id = "C16181", fx_groups_df = example_compound_fx_groups)
print(example_compound_vol$log_c)
```

This example compound has a volatility around `r round(example_compound_vol$log_c, 0)`. It is in the `r example_compound_vol$volatility` volatility category. 

### Estimate volatility of multiple chemical compounds used in a specified metabolic pathway usage

The `calc_pathway_vol` function will return a dataframe with volatility estimates for all of the compounds in the **map00361** pathway. 

```{r}
example_pathway_vol <- calc_pathway_vol("map00361", "data/")
print(example_pathway_vol[1,])
```

## Dataframe columns

The dataframe returned from the functions above contain the following fields, divided into categories below for clarity:

Basic compound information

- pathway: KEGG pathway identifier
- compound: KEGG compound identifier
- formula: compound chemical formula
- name: compound name
- mass: compound mass

The number of functional groups or atoms in the following classes

- carbons  
- ketones  
- aldehydes
- hydroxyl_groups 
- carbox_acids    
- peroxide 
- hydroperoxide   
- nitrate  
- nitro    
- carbon_dbl_bonds
- rings    
- rings_aromatic  
- phenol   
- nitrophenol     
- nitroester      
- ester    
- ether_alicyclic 
- ether_aromatic  
- amine_primary   
- amine_secondary 
- amine_tertiary  
- amine_aromatic  
- amines   
- amides   
- phosphoric_acid 
- phosphoric_ester
- sulfate  
- sulfonate
- thiol    
- carbothioester  
- oxygens  
- chlorines
- nitrogens
- sulfurs  
- phosphoruses    
- bromines 
- iodines  
- fluorines

Volatility calculation steps

- log_alpha: intermediate step
- log_Sum: intermediate step
- log_c: estimated volatility
- volatility: volatility category, where values less than 0 are "none", values between 0 and 2 are "moderate", and values above 2 are "high"

### References

Dan Tenenbaum and Bioconductor Package Maintainer (2021). "KEGGREST: Client-side REST access to the Kyoto Encyclopedia of Genes and Genomes (KEGG)". R package version 1.34.0.
  
Cao, Y., A. Charisi, L.-C. Cheng, T. Jiang, and T. Girke. 2008. "ChemmineR: A Compound Mining Framework for R." Bioinformatics 24 (15): 1733–4. https://doi.org/10.1093/bioinformatics/btn307.