---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# volcalc

<!-- badges: start -->
<!-- badges: end -->

The goal of volcalc is to automate calculating estimates of volatility for compounds. 

## Installing and loading package

Because the package's GitHub repo is private, you'll need to create a personal access token to install it. Get token by going to [https://github.com/settings/tokens](https://github.com/settings/tokens). Hit "Generate new token" button, give the token a name like "volcalc package", and make sure there is a check by the "repo" option before hitting green button at bottom of page. Copy the token, which is a random string of numbers and letters, on the next page and insert into the R command below. Do not share your token with anyone! 
You can install volcalc from [the GitHub repository](https://github.com/Meredith-Lab/volcalc) with:

``` r
# install.packages("devtools")
devtools::install_github("Meredith-Lab/volcalc", auth_token = "[token]")
```

You might be prompted to install some additional R packages, follow the instructions provided to do so. 

And then use the package with: 

```{r}
library(volcalc)
```
## Single compound usage

This is a basic example which shows you how to get a volatility estimate for an example compound *beta-2,3,4,5,6-Pentachlorocyclohexanol*. The KEGG compound identifier for the compound, as found on [the compound's KEGG page](https://www.genome.jp/dbget-bin/www_bget?C16181), is *C16181*. It is part of two molecular pathways; we will use *map00361* below. 

#### Single function approach

```{r}
calc_vol("map00361", "data", compound_id = "C16181")
```

This returns a dataframe with columns specifying general info about the compound, and the compound's calculated volatility and corresponding volatility category. The functional group counts underlying the volatility can be additionally returned with `return_fx_groups = TRUE`, and the intermediate calculation steps with `return_calc_steps = TRUE`. A list of all possible dataframe columns is included below. 

The compound can alternatively be specified with its chemical formula using the `compound_formula` argument instead of `compound_id` as in the example. 

#### Multiple function approach

This breaks the steps done by `calc_vol` into three parts: 1) download the compound's .mol file from KEGG, 2) count occurrences of different functional groups, and 3) estimate volatility. This calculation uses the SIMPOL approach (Prankow and Asher, 2008). 

```{r}
save_compound_mol("map00361", "data/", compound_id = "C16181")
example_compound_fx_groups <- get_fx_groups("C16181", "map00361", "data/")
example_compound_vol <- calc_vol("map00361", "data/", compound_id = "C16181", fx_groups_df = example_compound_fx_groups)
print(example_compound_vol$log_c)
```

This example compound has a volatility around `r round(example_compound_vol$log_c, 0)`. It is in the `r example_compound_vol$volatility` volatility category. 

## Multiple compounds from a pathway usage

A dataframe with volatility estimates for all compounds in a chosen pathway can be returned as below. 

```{r}
example_pathway_vol <- calc_pathway_vol("map00361", "data/")
print(example_pathway_vol[1,])
```

## Dataframe columns

Basic compound information

- pathway: KEGG pathway identifier
- compound: KEGG compound identifier
- formula: compound chemical formula
- name: compound name
- mass: compound mass

Counted functional groups and atoms

- carbons  
- ketones  
- aldehydes
- hydroxyl_groups 
- carbox_acids    
- peroxide 
- hydroperoxide   
- nitrate  
- nitro    
- carbon_dbl_bonds
- rings    
- rings_aromatic  
- phenol   
- nitrophenol     
- nitroester      
- ester    
- ether_alicyclic 
- ether_aromatic  
- amine_primary   
- amine_secondary 
- amine_tertiary  
- amine_aromatic  
- amines   
- amides   
- phosphoric_acid 
- phosphoric_ester
- sulfate  
- sulfonate
- thiol    
- carbothioester  
- oxygens  
- chlorines
- nitrogens
- sulfurs  
- phosphoruses    
- bromines 
- iodines  
- fluorines

Volatility calculation steps

- log_alpha: intermediate step
- log_Sum: intermediate step
- log_c: estimated volatility
- volatility: volatility category, where values less than 0 are "none", values between 0 and 2 are "moderate", and values above 2 are "high"
