---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# volcalc

<!-- badges: start -->

[![](https://github.com/Meredith-Lab/volcalc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Meredith-Lab/volcalc/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

## Overview

The goal of volcalc is to automate calculating estimates of volatility for chemical compounds.

Volatility can be estimated for most chemical compounds that are in the [KEGG](https://www.genome.jp/kegg/) database, using just the [KEGG unique identifier](https://www.genome.jp/kegg/compound/) for the compound of interest.
Alternatively, volatility can be estimated for multiple compounds that are in a [KEGG pathway](https://www.genome.jp/kegg/pathway.html).

## Installation

### Using without installing

You can use the `volcalc` package by using RStudio on a server [here](https://mybinder.org/v2/gh/Meredith-Lab/binder_volcalc/master?urlpath=rstudio).
This instance can be slow to launch.

The instance was generated using [Binder](https://mybinder.org/), which is an excellent free, open source tool to create custom computing environments.

To see an example of how to use `volcalc`, run the script `example_volcalc_usage.R` which is included in the server's file system.

### Installing locally

You can install `volcalc` from GitHub with

``` r
# install.packages("remotes")
remotes::install_github("Meredith-Lab/volcalc")
```

Or from r-universe with

``` r
install.packages('volcalc', repos = c('https://cct-datascience.r-universe.dev', 'https://cloud.r-project.org'))
```

Installation of `volcalc` requires the system library [OpenBabel](https://openbabel.org/) (it's a requirement of the `ChemmineOB` package, which `volcalc` depends on).
For macOS, this can be installed via homebrew by running the following shell command:

``` bash
brew install open-babel
```

For other installation options see the [OpenBabel documentation](https://openbabel.org/docs/dev/Installation/install.html).

### Loading package

Use the package with:

```{r}
library(volcalc)
```

## Single compound usage

This is a basic example which shows you how to get a volatility estimate for an example compound *beta-2,3,4,5,6-Pentachlorocyclohexanol*.
The KEGG compound identifier for the compound, as found on [the compound's KEGG page](https://www.genome.jp/dbget-bin/www_bget?C16181), is *C16181*.

#### Single function approach

```{r}
calc_vol(compound_id = "C16181")
```

This returns a dataframe with columns specifying general info about the compound, and the compound's calculated volatility and corresponding volatility category.
The functional group counts underlying the volatility can be additionally returned with `return_fx_groups = TRUE`, and the intermediate calculation steps with `return_calc_steps = TRUE`.
A list of all possible dataframe columns is included below.

There are other possible input arguments to the function.
The compound can alternatively be specified with its chemical formula using the `compound_formula` argument instead of `compound_id` as in the example.
The KEGG pathway that a compound is part of can be included with the `pathway_id` argument, which will generate a data subfolder for all compounds in that specified pathway.
You can specify where the compound files are downloaded by setting the desired relative path using `path = "path/to/folder"`; otherwise, the path will be in a `data` folder in the current directory.
If the underlying data file for a compound has already been downloaded in the specified path, it will not be downloaded again unless `redownload = TRUE`.

#### Multiple function approach

This breaks the steps done by `calc_vol` into three parts: 1) download the compound's .mol file from KEGG, 2) count occurrences of different functional groups, and 3) estimate volatility.
This calculation uses the SIMPOL approach[^1].

```{r}
save_compound_mol(compound_id = "C16181")
example_compound_fx_groups <- get_fx_groups(compound_id = "C16181")
example_compound_vol <- calc_vol(compound_id = "C16181", fx_groups_df = example_compound_fx_groups)
print(example_compound_vol$volatility)
```

This example compound has a volatility around `r round(example_compound_vol$volatility, 0)`.
It is in the `r example_compound_vol$category` volatility category.

Many of the arguments described for `calc_vol` can be used in these intermediate functions.
See function documentation for details.

## Multiple compounds from a pathway usage

A dataframe with volatility estimates for all compounds in a chosen pathway can be returned as below.

```{r}
example_pathway_vol <- calc_pathway_vol("map00361")
print(example_pathway_vol[1,])
```

## Dataframe columns

### Basic compound information

-   pathway: KEGG pathway identifier
-   compound: KEGG compound identifier
-   formula: compound chemical formula
-   name: compound name
-   mass: compound mass

### Counted functional groups and atoms

-   carbons
-   ketones
-   aldehydes
-   hydroxyl_groups
-   carbox_acids
-   peroxide
-   hydroperoxide
-   nitrate
-   nitro
-   carbon_dbl_bonds
-   rings
-   rings_aromatic
-   phenol
-   nitrophenol
-   nitroester
-   ester
-   ether_alicyclic
-   ether_aromatic
-   amine_primary
-   amine_secondary
-   amine_tertiary
-   amine_aromatic
-   amines
-   amides
-   phosphoric_acid
-   phosphoric_ester
-   sulfate
-   sulfonate
-   thiol
-   carbothioester
-   oxygens
-   chlorines
-   nitrogens
-   sulfurs
-   phosphoruses
-   bromines
-   iodines
-   fluorines

### Volatility calculation steps

-   log_alpha: intermediate step
-   log_Sum: intermediate step
-   volatility: estimated volatility
-   category: volatility category, where values less than 0 are "none", values between 0 and 2 are "moderate", and values above 2 are "high"

### Functional group details

| Functional group   | In manual? | Count method        | Coefficient | Coef source      |
|---------------|---------------|---------------|---------------|---------------|
| Carbons            | Y          | ChemmineR atomcount | -0.438      | ?                |
| Ketones            | Y          | ChemmineR groups    | -0.935      | Pankow & Asher   |
| Aldehydes          | Y          | ChemmineR groups    | -1.35       | Pankow & Asher   |
| Hydroxyl groups    | Y          | ChemmineR groups    | -2.23       | Pankow & Asher   |
| Carboxylic acids   | Y          | ChemmineR groups    | -3.58       | Pankow & Asher   |
| Peroxide           | Y          | SMARTS              | -0.368      | Pankow & Asher   |
| Hydroperoxide      | Y          | NA                  | -2.48       | Pankow & Asher   |
| Nitrate            | Y          | SMARTS              | -2.23       | Pankow & Asher   |
| Nitro              | Y          | SMARTS              | -2.15       | Pankow & Asher   |
| Carbon double bond | Y          | ChemmineR conMA     | -0.105      | Pankow & Asher   |
| Non-aromatic rings | Y          | ChemmineR rings     | --0.0104    | Pankow & Asher   |
| Aromatic rings     | Y          | ChemmineR rings     | -0.675      | Pankow & Asher   |
| Phenol             | Y          | SMARTS              | -2.14       | Pankow & Asher   |
| Nitrophenol        | Y          | NA                  | 0.0432      | Pankow & Asher   |
| Nitroester         | Y          | NA                  | -2.67       | Pankow & Asher   |
| Ester              | Y          | ChemmineR groups    | -1.20       | Pankow & Asher   |
| Ether (acyclic)    | Y          | NA                  | -0.683      | Pankow & Asher   |
| Ether (aromatic)   | Y          | NA                  | -1.03       | Pankow & Asher   |
| Amine primary      | Y          | ChemmineR groups    | -1.03       | Pankow & Asher   |
| Amine secondary    | Y          | ChemmineR groups    | -0.849      | Pankow & Asher   |
| Amine tertiary     | Y          | ChemmineR groups    | -0.608      | Pankow & Asher   |
| Amine aromatic     | Y          | ChemmineR rings     | -1.61       | Pankow & Asher   |
| Amine              | N          | SMARTS              | -2.23       | Same as nitrate  |
| Amide              | N          | SMARTS              | -2.23       | Same as nitrate  |
| Phosphoric acid    | N          | SMARTS              | -2.23       | Same as nitrate  |
| Phosphoric ester   | N          | SMARTS              | -2.23       | Same as nitrate  |
| Sulfate            | N          | SMARTS              | -2.23       | Same as nitrate  |
| Sulfonate          | N          | SMARTS              | -2.23       | Same as nitrate  |
| Thiol              | N          | SMARTS              | -2.23       | Same as hydroxyl |
| Carbothioester     | N          | SMARTS              | -1.20       | Same as ester    |

## How to contribute

We appreciate many kinds of feedback and contributions to this R package.
If you find a bug, are interested in an additional feature, or have made improvements to the package that you want to share, feel free to file an [issue](https://github.com/Meredith-Lab/volcalc/issues/new) in this GitHub repo.

## How to cite

If you use this package in your published work, please cite it using the reference below:

> Meredith, L.K., Riemer, K., Geffre, P., Honeker, L., Krechmer, J., Graves, K., Tfaily, M., and Ledford, S.K. Automating methods for estimating metabolite volatility.
> In prep.

### References

[^1]: Pankow, J.F., Asher, W.E., 2008. SIMPOL.1: a simple group contribution method for predicting vapor pressures and enthalpies of vaporization of multifunctional organic compounds. Atmos. Chem. Phys. https://doi.org/10.5194/acp-8-2773-2008

