---
title: "Introduction to volcalc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to volcalc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(volcalc)
```

The primary function in `volcalc` is `calc_vol()`.  It accepts either a path to a .mol file or a SMILES string.  There are a few example .mol files included in the `volcalc` installation.  The paths to these can be found with `mol_example()`

### Basic usage with .mol files

```{r}
#list possible mol files
mol_example()

#get path to specific file
mol_path <- mol_example("C16181.mol")
mol_path
```

The default output of `calc_vol()` includes a relative volatility index, `rvi` which is equivalent to $\textrm{log}_{10}C^\ast$.  It also includes a RVI category for clean air.

```{r}
calc_vol(mol_path)
```

### Specify environment

Specifying `environment` only alters the RVI category by using different RVI cutoffs for non-volatile, low, moderate, and high volatility.

```{r}
calc_vol(mol_path, environment = "soil")
```

### Return intermediate steps

`calc_vol()` uses a modified version of the SIMPOL.1 method by default which is a group contribution method.  You can have `calc_vol()` return the counts of functional groups and other molecular properties (which is useful for validation) with `return_fx_groups = TRUE`.  See `?get_fx_groups()` for more information about these additional columns.

```{r}
calc_vol(mol_path, return_fx_groups = TRUE)
```

The SIMPOL.1 method calculates $\textrm{log}_{10} P_{\textrm{L},i}^\circ(T)$, which is used by `calc_vol()` to calculate RVI. To see the intermediate calculations, use `return_calc_steps = TRUE`.

```{r}
calc_vol(mol_path, return_calc_steps = TRUE)
```

### Use with SMILES

Finally, all of this can be done using SMILES strings rather than .mol files with `from = "smiles"`.  Backslash, `\` is a valid SMILES character, but isn't a valid character in R and must be "escaped" as `\\`.

```{r}
## This will error even though the SMILES is correct
# calc_vol("CC/C=C\C[C@@H]1[C@H](CCC1=O)CC(=O)O", from = "smiles")

# To solve this, escape \C as \\C
calc_vol("CC/C=C\\C[C@@H]1[C@H](CCC1=O)CC(=O)O", from = "smiles")
```

